---
title: "PopHelper analysis of FastStructure results"
author: "Natalie Forsdick"
date: "17/09/2024"
header-includes:
    - \usepackage{setspace}\doublespacing
output:
  html_document: 
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/path/to/Weta-GBS/2024-11-01-Results/FastStructure", echo = TRUE,
                     dev='tiff', fig.path='./figures/', dpi=300, units="in", fig.height=7, fig.width=10, 
                     cache=FALSE)
#rm(list = ls(all.names = TRUE))
getwd()

```

```{r packages, include=FALSE}
# Commented code is what had to be set up initially to load pophelper:
#remove.packages('ggplot2')
#install devtools package from CRAN
#install.packages('devtools',dependencies=T)
#library(devtools) # may need to run this but will check first

#install pophelper package from GitHub:
#install_github('royfrancis/pophelper', force=TRUE)

#load library for use
library('pophelper')
packageDescription("pophelper", fields="Version")
R.version
#library(RColorBrewer)
library(tidyr)

```

```{r colours, include=FALSE}
# Set colours for plots
#greenbluegrey <- c("#6F808C", "#70CFAE", "#29AB7D","#CECECE", "#32D199","#28B0D1","#143A8B",  "#F0A156", "#5B88C1","#C05E23", "#99B5DC", "#F7B969")
greenbluegrey <- c("#009cbd", "#de7c00", "#b7db57", "#64a70b",  "#00c1d5", "#ebb700", "#32D199", "#5B88C1", "purple", "#C05E23")

oranges <- c("#de7c00","#ebb700","#C05E23")
```

Here we use the `R` package `PopHelper` to plot the outputs of FastStructure. 

Prior to importing our data, we have to do some pre-processing to ensure iteration numbers are appended to file names, and that all *Q* files are in the same directory. We also want to extract our CVs so we can identify the most appropriate *k*-values. This pre-processing uses the command line and is described in Step 7 of 'Streamlined_GBS_pipeline.md'.

## Assess CV to determine the most appropriate number of clusters

This requires all CVs to be extracted and sorted by *k*, in a single file `admixture_CV_sorted.txt`. 

```{r CV_values, echo = FALSE}
# Import data file
cvfile <- read.table(file="faststr_cv.txt",header= FALSE, sep= "\t", stringsAsFactors = FALSE)

# name headers
names(cvfile) <- c("k","CV")

head(cvfile)
```

```{r CV_values_means, echo = FALSE}
# To generate group-specific means we can use aggregate function
cvfile_means <- aggregate(cvfile[,2], list(cvfile$k), mean)
cvfile_means
```

```{r CV_values_plot, fig.height=4, fig.width=6, echo = FALSE}
# can we plot without having a manual step of adding means to the file?
library(ggplot2)
library(forcats)
# using forcats function 'fct_inseq' to order x axis numerically after grouping as factors

# Then plot of CV by k.
tiff("./figures/Figure3.2.tiff", compression = "lzw", units="in", width=5, height=4.5, res=400)
ggplot(cvfile, aes(fct_inseq(factor(k), ordered = NA), CV, group = 1)) +
  geom_point(colour="grey") +
  geom_line(stat = "summary", fun = "mean", linewidth=1)+
  theme_bw() +
  labs(x="K", y="CV error")
dev.off()
```


Now we can identify our most appropriate *K* value using this plot. For this first SNP set, k = 2 is most appropriate. 

Hm, but somewhere else you must have the code to do the actual Evanno method plots for all.

Next we can import our *Q* files to `PopHelper`, and begin plotting.
First, make a list of files to include.

```{r make_Qlist}
q1<- list.files(path="./", pattern = c("*.meanQ"), full.names=T)

# Check files are correct:
head(q1)
```

Create a list

```{r make_Qlist2}
alist <- readQ(files=q1, indlabfromfile=T)
as.qlist(alist)

```

We need a txt file containing the population information for individuals 'admix_filt_indiv.txt'. 

```{r confirm_Qlist2}
# Had some issues later with k or ind labels not set properly, so using this to try and solve that problem.
q2 <- as.qlist(alist)
#head(q2)
is.qlist(q2)

popcode <- read.delim("../weta.snpmiss60.mac3.thin_grouplab.txt", header=T, stringsAsFactors=F)
#head(popcode)

# Tabulate Q files, and make a table
tr1 <- tabulateQ(q2, writetable=F)
#tr1
# Summarise Q files, collapsing replicate runs
sr1 <- summariseQ(tr1)
#head(sr1)

#summariseQ(tr1, writetable=TRUE)
```

Next create an initial unaligned plot, just to check that everything imported and is running as expected.

```{r test_plot, eval=FALSE, fig.height=10, fig.width=6, include=FALSE}
p1 <- plotQ(q2[c(1:10)], imgoutput="join", returnplot=T, exportplot=T, outputfilename="test_plot", imgtype="tiff", basesize=11, exportpath="./figures/")

plotQ(q2[c(1:10)],imgoutput="join",showindlab=T,useindlab=T,height=7,width=70,grplabangle=0,ordergrp=T,basesize=11,returnplot=T, exportplot=T,  outputfilename="test_plot",imgtype="tiff",exportpath="./figures/")
p1
```



Now this works for a set with the same *K* value - what happens if we do it for all data?

```{r align_all, fig.height=16, fig.width=8, echo = FALSE}
alist2 <- alignK(q2)
```


```{r align_all2, eval=FALSE, fig.height=11, fig.width=8, include=FALSE}
plotQ(alist2, imgoutput="join", returnplot=T, showindlab=F, grplab=popcode, exportplot=T, outputfilename="align_all2", imgtype="tiff", basesize=12, exportpath="./figures/",
      clustercol=greenbluegrey,
      height=1, indlabsize=2, indlabheight=0.08, indlabspacer=-1)
```

We then convert to CLUMPP format. 

```{r clumpp_output, eval=FALSE, include=FALSE}
clumppExport(qlist=q2, prefix="Faststr_filt",exportpath=getwd())
```

Ran these through CLUMPP. This should provide support for the most appropriate *K* value determined above. 

```{bash eval=F}
for file in Faststr_filt_K*/
do
  cd $file
  "/path/to/CLUMPP_Windows.1.1.2/CLUMPP.exe" paramfile
  cd ../
done
```


Results:

| K  | 2 | 3 | 4 | 5 | 
| H' | 1 | 0.4565 | 0.3881 | 0.4121 | 

So we are looking for the K value that gives the highest H' value. 

If required, you can plot your merged files for comparison. 

```{r clumpp_results, echo=FALSE, fig.height=5, fig.width=10}
aligned <- readQ("./Faststr_filt_K2/Faststr_filt_K2-combined-aligned.txt")
```

```{r clumpp_aligned_plot, echo=FALSE, fig.height=20, fig.width=10}
plotQ(aligned, imgoutput="join", grplab=popcode, subsetgrp=c("MI","MGWSR"), selgrp="Population", ordergrp=T, clustercol=greenbluegrey, outputfilename="clumpped_k2", exportpath=getwd())
```

```{r clumpp_merged, echo=FALSE}

merged2 <- readQ("./Faststr_filt_K2/Faststr_filt_K2-combined-merged.txt")
merged3 <- readQ("./Faststr_filt_K3/Faststr_filt_K3-combined-merged.txt")

```

```{r clumpp_merged_plot1, echo=FALSE, fig.height=5, fig.width=10}
popcode1 <- read.delim("../weta.snpmiss60.mac3.thin-pop.txt", header=F, stringsAsFactors=F)
plotQ(merged2, grplab=popcode1, subsetgrp=c("MI","MGWSR"), selgrp="V2", ordergrp=T,
      showindlab = F,
      showlegend=F, exportplot=T, outputfilename="CLUMPP_merged_k2-pop",imgtype="tiff", exportpath = "./figures/", basesize=8, 
      grplabangle = 0, grplabsize = 1.4, splabsize = 3, divtype = 1, divsize = 0.2, panelspacer = 0.06, grplabspacer = -0.2, linesize = 0.3,
      grplabheight = 0.2, 
             barbordercolour="white", barbordersize=0.01, clustercol=oranges, height=4, dpi=500)

plotQ(merged3, grplab=popcode1, subsetgrp=c("MI","MGWSR"), selgrp="V2", ordergrp=T,
      showindlab = F,
      showlegend=F, exportplot=T, outputfilename="CLUMPP_merged_k3-pop", imgtype="tiff", exportpath ="./figures/", basesize=8, 
      grplabangle = 0, grplabsize = 1.4, splabsize = 3, divtype = 1, divsize = 0.2, panelspacer = 0.06, grplabspacer = -0.2, linesize = 0.3,
      grplabheight = 0.2, 
             barbordercolour="white", barbordersize=0.01, clustercol=oranges, height=4, dpi=500)
```


```{r clumpp_merged_plot2, echo=FALSE, fig.height=5, fig.width=10}

plotQ(merged2, grplab=popcode,subsetgrp=c("MI","MGWSR"), selgrp="Population", ordergrp=T,
      showindlab = F,
      showlegend=F, exportplot=T, outputfilename="CLUMPP_merged_k2",imgtype="tiff", exportpath = "./figures/", basesize=8, 
      grplabangle = 35, grplabsize = 1, splabsize = 3, divtype = 1, divsize = 0.2, panelspacer = 0.06, grplabspacer = -0.2, linesize = 0.3,
      grplabheight = 0.2, 
             barbordercolour="white", barbordersize=0.01, clustercol=oranges, height=4, dpi=500)

plotQ(merged3, grplab=popcode, subsetgrp=c("MI","MGWSR"), selgrp="Population", ordergrp=T,
      showindlab = F,
      showlegend=F, exportplot=T, outputfilename="CLUMPP_merged_k3", imgtype="tiff", exportpath ="./figures/", basesize=8, 
      grplabangle = 35, grplabsize = 1, splabsize = 3, divtype = 1, divsize = 0.2, panelspacer = 0.06, grplabspacer = -0.2, linesize = 0.3,
      grplabheight = 0.2, 
             barbordercolour="white", barbordersize=0.01, clustercol=oranges, height=4, dpi=500)

```

```{r clumpp_merged_sex_col, echo=FALSE, fig.height=5, fig.width=10}
# had to change image output format to png, as can't seem to read a tiff into Rmarkdown when building supplementary

plotQ(merged2, grplab=popcode,subsetgrp=c("M","F"), selgrp="Sex", ordergrp=T,
      showindlab = F,
      showlegend=F, exportplot=T, outputfilename="CLUMPP_merged_k2_sex",imgtype="png", exportpath = "./figures/", basesize=8, 
      grplabangle = 35, grplabsize = 1, splabsize = 3, divtype = 1, divsize = 0.2, panelspacer = 0.06, grplabspacer = -0.2, linesize = 0.3,
      grplabheight = 0.2, 
             barbordercolour="white", barbordersize=0.01, clustercol=oranges, height=4, dpi=500)

plotQ(merged3, grplab=popcode,subsetgrp=c("M","F"), selgrp="Sex", ordergrp=T,
      showindlab = F,
      showlegend=F, exportplot=T, outputfilename="CLUMPP_merged_k3_sex",imgtype="png", exportpath = "./figures/", basesize=8, 
      grplabangle = 35, grplabsize = 1, splabsize = 3, divtype = 1, divsize = 0.2, panelspacer = 0.06, grplabspacer = -0.2, linesize = 0.3,
      grplabheight = 0.2, 
             barbordercolour="white", barbordersize=0.01, clustercol=oranges, height=4, dpi=500)

plotQ(merged2, grplab=popcode, subsetgrp=c("Dark","Light", "Yellow"), selgrp="Colour", ordergrp=T,
      showindlab = F,
      showlegend=F, exportplot=T, outputfilename="CLUMPP_merged_k2_col", imgtype="png", exportpath ="./figures/", basesize=8, 
      grplabangle = 35, grplabsize = 1, splabsize = 3, divtype = 1, divsize = 0.2, panelspacer = 0.06, grplabspacer = -0.2, linesize = 0.3,
      grplabheight = 0.2, 
             barbordercolour="white", barbordersize=0.01, clustercol=oranges, height=4, dpi=500)

plotQ(merged3, grplab=popcode, subsetgrp=c("Dark","Light", "Yellow"), selgrp="Colour", ordergrp=T,
      showindlab = F,
      showlegend=F, exportplot=T, outputfilename="CLUMPP_merged_k3_col", imgtype="png", exportpath ="./figures/", basesize=8, 
      grplabangle = 35, grplabsize = 1, splabsize = 3, divtype = 1, divsize = 0.2, panelspacer = 0.06, grplabspacer = -0.2, linesize = 0.3,
      grplabheight = 0.2, 
             barbordercolour="white", barbordersize=0.01, clustercol=oranges, height=4, dpi=500)
```

