---
title: "SNPRelate_test"
author: "Nat Forsdick"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    keep_md: TRUE
---

Run on Linux system

```{r setup, include=T}
knitr::opts_knit$set(dev = "tiff", echo = TRUE,ppi=300,units="in", out.height="7", out.width="9", cache = TRUE)

getwd()
```

## SNPRelate for PCA

```{r install_packages, include=F, eval=F}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("gdsfmt")
BiocManager::install("SNPRelate")
BiocManager::install("gdsfmt")
BiocManager::install("MASS")
install.packages("pals")
install.packages("here")
here::here()
```

```{r libraries, echo=FALSE}
#install.packages("SNPRelate")

library(gdsfmt)
library(SNPRelate)
library(MASS)
library(RColorBrewer)
#library("Manu")
library(pals)
library('gridExtra')
library(vcfR)
```

```{r session, echo=FALSE}
sessionInfo()

citation("gdsfmt")
citation("SNPRelate")
citation("MASS")
```

```{r colours}
#Titipounamu <- get_pal("Titipounamu")
#palette(Titipounamu)
# Create a gradient of colours between the selected colours
#selected_colours <- get_pal("Titipounamu")[c(3,5,6,2)]
selected_colours <- c("#009cbd", "#ebb700", "#de7c00")
color.gradient <- function(x, colors=selected_colours, colsteps=100) {
  return( colorRampPalette(colors) (colsteps) [ findInterval(x,
                seq(min(x),max(x), length.out=colsteps)) ] )
}
x <- c(1:50)

#selected_colours2 <- get_pal("Titipounamu")[c(1,5)]
selected_colours2 <- c("#009cbd", "#de7c00")
color.gradient2 <- function(x, colors=selected_colours2, colsteps=100) {
  return( colorRampPalette(colors) (colsteps) [ findInterval(y,
                seq(min(y),max(y), length.out=colsteps)) ] )
}
y <- c(1:50)

mwlrcols <- c("#00c1d5","#64a70b", "#009cbd",  "#b7db57", "#de7c00", "#ebb700")
oranges <- c("#de7c00","#ebb700","#C05E23")
```

```{r import_vcfs}
# check vcf is not zipped
vcf.fn <- "/path/to/Weta-GBS/2024-11-01-Results/weta.snpmiss60.mac3.thin.vcf"

#file.size(vcf.fn)
#vcf_test1 <- read.vcfR(vcf.fn)
snpgdsVCF2GDS("/path/to/Weta-GBS/2024-11-01-Results/weta.snpmiss60.mac3.thin.vcf", "test.gds", method="biallelic.only")

snpgdsSummary("test.gds")

```

```{r sanity_check}
# Open the GDS file
genofile <- snpgdsOpen("test.gds")

head(genofile)
```

```{r get_pop}
# Get population information
#pop_code <- scan("/path/to/Weta-GBS/2024-11-01-Results/weta.snpmiss60.mac3.thin-pop.txt",
#                 what=character())
pop_code <- read.delim("/path/to/Weta-GBS/2024-11-01-Results/weta.snpmiss60.mac3.thin-pop.txt", sep="\t", header=F)
table(pop_code)

# Display the first six values
head(pop_code)

# Get sample id
sample.id <- read.gdsn(index.gdsn(genofile, "sample.id"))
sample.id
# assume the order of sample IDs is as the same as population codes
head(cbind(sample.id, pop_code$V2))
```


```{r PCA}
population <- as.factor(pop_code$V2)
pca <- snpgdsPCA(genofile, num.thread=4,autosome.only=F)

# variance proportion (%)
pc.percent <- pca$varprop*100
head(round(pc.percent, 2))

# Make the data frame
tab <- data.frame(sample.id = pca$sample.id,
    pop = factor(pop_code$V2)[match(pca$sample.id, sample.id)],
    EV1 = pca$eigenvect[,1],    # the first eigenvector
    EV2 = pca$eigenvect[,2],    # the second eigenvector
    stringsAsFactors = FALSE)
head(tab)

# Draw
legend.cols = as.factor(oranges[tab$pop])

plot(tab$EV2, tab$EV1, pch=16, col=(oranges)[as.integer(tab$pop)], 
     xlab="Eigenvector 2 (2.4%)", ylab="Eigenvector 1 (3.0%)")
legend("bottomright", legend=levels(tab$pop), 
       pch=16, col=levels(legend.cols))

tiff("Figure2.tiff", compression = "lzw", units="in", width=6, height=4, res=400)
plot(tab$EV2, tab$EV1, pch=16, col=(oranges)[as.integer(tab$pop)], 
     xlab="Eigenvector 2 (2.4%)", ylab="Eigenvector 1 (3.0%)")
legend("bottomright", legend=levels(tab$pop), 
       pch=16, col=levels(legend.cols))
dev.off()

lbls <- paste("PC", 1:5, "\n", format(pc.percent[1:5], 
                                      digits=2), "%", sep="")
pairs(pca$eigenvect[,1:5], pch=16, col=(oranges)[as.integer(tab$pop)], labels=lbls)
legend("topleft", legend=levels(tab$pop), 
       pch=16, col=levels(legend.cols))

lbls <- paste("PC", 1:2, "\n", format(pc.percent[1:2], 
                                      digits=2), "%", sep="")
pairs(pca$eigenvect[,1:2], pch=16, col=(oranges)[as.integer(tab$pop)], labels=lbls)
legend("topleft", legend=levels(tab$pop), 
       pch=16, col=levels(legend.cols))

tiff("Figure2.2.tiff", compression = "lzw", units="in", width=6, height=4, res=400)
pairs(pca$eigenvect[,1:2], pch=16, col=(oranges)[as.integer(tab$pop)], labels=lbls)
legend("topleft", legend=levels(tab$pop), 
       pch=16, col=levels(legend.cols))
dev.off()
```

Parallel coordinates plot for the top principal components:
```{r top_PCs}
datpop <- factor(pop_code$V2)[match(pca$sample.id, sample.id)]

levelcol <- levels(datpop)
parcoord(pca$eigenvect[,1:16], col=oranges[datpop])
legend("topleft", legend=levels(datpop), 
       pch=16, col=(oranges)[as.factor(levelcol)])

```

To calculate the SNP correlations between eigenvectors and SNP genotypes:

```{r get_corr}
palette(alphabet2())
# Get chromosome index
chr <- read.gdsn(index.gdsn(genofile, "snp.chromosome"))
CORR <- snpgdsPCACorr(pca, genofile, eig.which=1:4)

savepar <- par(mfrow=c(2,1), mai=c(0.45, 0.55, 0.1, 0.25))

for (i in 1:2)
{
    plot(abs(CORR$snpcorr[i,]), ylim=c(0,1), xlab="",
         ylab=paste("PC", i),
         col=1:length(chr), pch="+")
}
```

Can also do Weir-Cockerham Fst
```{r fst}
flag <- pop_code$V2 %in% c("MI", "SR")
samp.sel <- sample.id[flag]
pop.sel <- pop_code$V2[flag]
v <- snpgdsFst(genofile, sample.id=samp.sel, population=as.factor(pop.sel),
    method="W&C84", autosome.only=F)

# Weir and Cockerham weighted Fst estimate
v$Fst 
# Weir and Cockerham mean Fst estimate
v$MeanFst    
summary(v$FstSNP)
```


```{r MI_relatedness,include=F,eval=T,echo=T}
MI.id <- sample.id[pop_code$V2 == "MI"]
ibd <- snpgdsIBDMoM(genofile, sample.id=MI.id,
    maf=0.05, missing.rate=0.05, num.thread=4, autosome.only=F)

ibd.coeff <- snpgdsIBDSelection(ibd)
head(ibd.coeff)

plot(ibd.coeff$k0, ibd.coeff$k1, xlim=c(0,1), ylim=c(0,1),
    xlab="k0", ylab="k1", main="MI samples (MoM)")
lines(c(0,1), c(1,0), col="red", lty=2)
```


```{r SR_relatedness,include=F,eval=T,echo=F}
SR.id <- sample.id[pop_code$V2 == "SR"]
ibd <- snpgdsIBDMoM(genofile, sample.id=SR.id,
    maf=0.05, missing.rate=0.05, num.thread=4, autosome.only=F)

ibd.coeff <- snpgdsIBDSelection(ibd)
head(ibd.coeff)

plot(ibd.coeff$k0, ibd.coeff$k1, xlim=c(0,1), ylim=c(0,1),
    xlab="k0", ylab="k1", main="MGWSR samples (MoM)")
lines(c(0,1), c(1,0), col="red", lty=2)
```

Now we can assess relatedness using a method-of-moments approach, first for MI:

```{r MI_relatedness_nofilt}
MI.id <- sample.id[pop_code$V2 == "MI"]
ibd <- snpgdsIBDMoM(genofile, sample.id=MI.id,
     num.thread=4, autosome.only=F)

ibd.coeff <- snpgdsIBDSelection(ibd)
head(ibd.coeff)

plot(ibd.coeff$k0, ibd.coeff$k1, xlim=c(0,1), ylim=c(0,1),
    xlab="k0", ylab="k1", main="MI samples (MoM)")
lines(c(0,1), c(1,0), col="red", lty=2)
```

And then for SR:

```{r SR_relatedness_nofilt}
SR.id <- sample.id[pop_code$V2 == "SR"]
ibd <- snpgdsIBDMoM(genofile, sample.id=SR.id,
    num.thread=4, autosome.only=F)

ibd.coeff <- snpgdsIBDSelection(ibd)
head(ibd.coeff)

plot(ibd.coeff$k0, ibd.coeff$k1, xlim=c(0,1), ylim=c(0,1),
    xlab="k0", ylab="k1", main="MGWSR samples (MoM)")
lines(c(0,1), c(1,0), col="red", lty=2)
```
We can then look at all individuals together:

```{r all_relatedness_nofilt}
ibd <- snpgdsIBDMoM(genofile, sample.id=sample.id,
    num.thread=4, autosome.only=F)

ibd.coeff <- snpgdsIBDSelection(ibd)
head(ibd.coeff)

plot(ibd.coeff$k0, ibd.coeff$k1, xlim=c(0,1), ylim=c(0,1),
    xlab="k0", ylab="k1", main="All samples (MoM)")
lines(c(0,1), c(1,0), col="red", lty=2)
```

Then we can look at identity-by-state:

```{r Identity-by-state}
ibs <- snpgdsIBS(genofile, num.thread=2, autosome.only=F)

pop.idx <- order(pop_code$V2)

image(ibs$ibs[pop.idx, pop.idx], col=color.gradient(y))
#image(ibs$ibs[pop.idx, pop.idx], col=terrain.colors(16))
```

To perform multidimensional scaling analysis on the n×n matrix of genome-wide IBS pairwise distances:

```{r mdsa}
#palette(Titipounamu)
loc <- cmdscale(1 - ibs$ibs, k = 2)
x <- loc[, 1]; y <- loc[, 2]
#popn <- oranges[as.factor(pop_code)]
popn <- as.factor(pop_code$V2)
plot(x, y, col=oranges[popn], xlab = "", ylab = "", pch=16,
    main = "Multidimensional Scaling Analysis (IBS)")
legend("topleft", legend=levels(popn), col=oranges, pch=16, text.col="black")

```
To perform cluster analysis on the n×n matrix of genome-wide IBS pairwise distances, and determine the groups by a permutation score:

```{r cluster}

set.seed(100)
ibs.hc <- snpgdsHCluster(snpgdsIBS(genofile,
                                   num.thread=2,autosome.only = F))

# Determine groups of individuals automatically
rv <- snpgdsCutTree(ibs.hc)

plot(rv$dendrogram, leaflab="none", main="HapMap Phase II")

table(rv$samp.group)

# Determine groups of individuals by population information
rv2 <- snpgdsCutTree(ibs.hc, samp.group=as.factor(pop_code$V2))

nodeCols <- as.factor(oranges[popn])
GROUP <- rv2$samp.group
cols <- levels(legend.cols)
col_GROUP <- cols[GROUP]
col_GROUP
col_GROUP <- col_GROUP[order.dendrogram(rv2$dendrogra)] 

library(dendextend)
library(colorspace)

dend <- rv2$dendrogram
#dend <- dend %>%
#  set("leaves_col", col_GROUP) %>%
#  plot(leaflab="none", ylab="Individual dissimilarity")
##  legend("bottomright", legend=levels(popn), col=levels(legend.cols), 
#       pch=19, ncol=4)
#dend

tiff("Figure3.tiff", compression = "lzw", units="in", width=6, height=4, res=400)

# Make a plot
dend %>%
  set("leaves_col", col_GROUP) %>%
  plot(leaflab="none", ylab="Individual dissimilarity")
  legend("bottomright", legend=levels(popn), col=levels(legend.cols), 
       pch=19, ncol=4, cex=0.7)

# Close device
dev.off()

#plot(rv2$dendrogram, fill=col_GROUP, leaflab="none",  main="HapMap Phase II")
#legend("bottomright", legend=levels(popn), col=levels(legend.cols), 
#       pch=19, ncol=4)

# add indiv labels
#plot(rv2$dendrogram, col=mwlrcols, leaflab="perpendicular",  main="HapMap Phase II")
#legend("bottomright", legend=levels(popn), col=levels(legend.cols),
#       pch=19, ncol=4)

dend <- rv2$dendrogram
tiff("Figure3.2.tiff", compression = "lzw", units="in", width=7, height=4, res=400)
dend %>%
  set("leaves_col", col_GROUP) %>%
  plot(leaflab="perpendicular", ylab="Individual dissimilarity")
  legend("bottomright", legend=levels(popn), col=levels(legend.cols), 
       pch=19, ncol=4, cex=0.6)

dev.off()

tiff("Figure3.2b.tiff", compression = "lzw", units="in", width=7, height=4, res=400)
dend %>%
  set("leaves_col", col_GROUP) %>%
  plot(leaflab="none", ylab="Individual dissimilarity")
  legend("bottomright", legend=levels(popn), col=levels(legend.cols), 
       pch=19, ncol=4, cex=0.6)

dev.off()
```


```{r close, echo=F}
# Close the GDS file
snpgdsClose(genofile)
```
